sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-27 doc/paper.Rmd
Edit

An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`).
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometers:

@@

We start by creating a mesh object that contains matrices to apply the SPDE approach:

@@

Here, `cutoff` defines the minimum allowed distance between points in the units of `X` and `Y` (km).
Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`. 
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)a).

@@

Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models.
We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method (omitted for space):

@@

We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we want to predict.
The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.

@@

We could extend our model to a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1):

@@

We could generate a standardized index from that fit on a 2x2 km grid:

@@

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter. 
The reasons for the interannual variation in the number of individuals migrating south are still not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution. 
We fit counts observed in North America during annual Christmas Bird Counts (**REF**) using a negative binomial ('NB2') distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with NAO (**see appendix for more detail**):

@@

We found a weak positive effect of annual mean NAO on overall counts, but also a southeast to northwest gradient in the intensity of this effect.
This result would be consistent with owls closer to the Atlantic coast and those migrating the furthest south being the most effected by NAO (Fig. \@ref(fig:owl-nao)).

(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020. 
Circles represent all count locations and are their area is scaled to the mean number of owls observed per year (range: 0 to 8).

>>>>

An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`).
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.

@@

We start by creating a mesh object that contains matrices to apply the SPDE approach.

@@

Here, `cutoff` defines the minimum allowed distance between points in the units of `X` and `Y` (km).
Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`. 
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)A).

@@

Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models.
We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method (omitted for space).

@@

We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we want to predict.
The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)B--D.

@@

We could extend our model to a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1).

@@

We could generate a standardized index from that fit on a 2x2 km grid by supplying the `area` per grid cell and telling `predict.sdmTMB()` to return a necessary TMB object (Fig. \@ref(fig:pcod-fig)E). `get_index()` sums the biomass across grid cells to derive an area-weighted index of biomass independent of sampling locations.

@@

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter. 
The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution. 
We fit counts observed in North America during annual Christmas Bird Counts (**REF**) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (**see appendix for more detail**):

@@

We found a weak average positive effect of annual mean NAO on overall counts, but a marked southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao)).
This result would be consistent with owls closer to the Atlantic coast and those migrating the furthest south being the most effected by NAO.

(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020. 
Circles represent all count locations and are their area is scaled to the mean number of owls observed per year (range: 0 to 8). 
The effect is multiplicative on owl count.
